Identifying timing of shifts in transboundary fish stocks to support climate resilient fisheries management

### 2021-03-10 Scripts/Manuscript.Rmd
updated manuscript for submission for gdrive

---
title: Early range shifts in transboundary fish stocks challenge their management under climate change
# subtitle: Changes in sharing of transboundary fisheries under climate change
author: 'Authors: Juliano Palacios-Abrantes^1*^, Thomas Frölicher^2,3^, Gabriel Reygondeau^1^, U. Rashid Sumaila^1^, Alessandro Tagliabue^4^, Colette Wabnitz^1,5^, and William W.L. Cheung^1^'
date: 'Affiliations: ^1^Institute for the Oceans and Fisheries, University of British Columbia, Vancouver, Canada. ^2^Climate and Environmental Physics, Physics Institute, University of Bern. ^3^Oeschger Centre for Climate Change Research, University of Bern, Switzerland. ^4^School of Environmental Sciences, University of Liverpool. ^5^Stanford Center for Ocean Solutions, Stanford University, Stanford, United States'
output:
  # bookdown::word_document2
    reference_docx: ../References/template.docx
  # pdf_document: default
bibliography: ../References/ToE_Ref.bib
csl: ../References/science.csl
header-includes:
- \usepackage{amsmath}
- \usepackage{float} # Use the 'float' package
- \floatplacement{figure}{H} # Make every figure with caption = h
---

@@

<!-- * corresponding author: j.palacios@oceans.ubc.ca -->

<!-- # Abstract (150  words) -->

@@

Climate change shifts in marine species distributions is affecting the current frameworks emplaced to manage species shared fisheries. Here, we rely on a modeling framework with multiple ensemble members to identify the time frame by which shared stocks distribution will change from historical ranges, and the change in the proportion of the distrirbution that each EEZ has under climate change. Our results suggest that 60% of shared species have emerged since 2006 in 87% of the world EEZs. By 2075 all EEZs will have an emerging species and by 2097 and all species will have emerged in at least one EEZ. Average changes in the ratio of species-share between neighboring EEZs will fluctuate between 1 and 10% with some exceeding 25% by 2030 relative to today’s ratio. Our results suggest that most changes will happen before 2030 urging policy to be re-formulated to account for the possible implications of these changes.

## Introduction

Over the last century, human activities have altered the physical and biogeochemical conditions of the ocean, including warming, acidification and reducing oxygen content [@IPCC:2019tn]. Distributions of marine species are closely related to the environment and the species’ preferences to environmental conditions (e.g., temperature, oxygen, salinity) [@Hutchinson:1957nt; @Nelson:2016kk]. As a result of climate change, many marine species have changed their distributions towards higher latitude, deeper water or following local temperature gradients [@Poloczanska:2016kk]. Biogeography of marine species is projected to continue to shift as ocean conditions are changing in the $21^{st}$ century under climate change [@Cheung:2010dt], impacting fisheries production and compromising our capacity to reach international sustainability goals such as Sustainable Development Goal 14 - life below water [@UnitedNations2018; @Barange:2014kr; @Singh:2017ds]. The projected risks of impacts can be reduced through improving the effectiveness of current fisheries managemen [@Gaines:2018sg; @Cheung:2018dh], including the fisheries management of species that cross international borders, i.e., shared stocks [@Gaines:2018sg; @Pinsky:2018cb].

The concept of shared stocks was developed following the ratification of the United Nations Convention on the Law of the Sea (UNCLOS) and the claiming of Exclusive Economic Zones (EEZs) by States [@UnitedNations:1986tl]. As defined by the United Nations’ Food and Agriculture Organization (FAO), shared stocks can be classified into four non-exclusive categories: (*i*) transboundary stocks, those that cross neighboring EEZs; (*ii*) straddling stocks, that, in addition to neighboring EEZs, also visit the adjacent high seas; (*iii*) highly migratory stocks, mainly tunas and bill-fishes, that migrate across vast oceanic regions including both the high seas and EEZs; and finally (*iV*) discrete stocks that are only present in the high seas [@Munro:2004th]. This study focuses on transboundary stocks exploited by fisheries operating within EEZs. While countries are responsible for the management of stocks within their EEZs, under UNCLOS, States are encouraged to cooperate when stocks are shared [@UnitedNations:1986tl]. A recent study estimates that, globally, there are 633 transboundary fish species representing 67% or the identified fished taxa, and yielding an annual average of 48.5 million tonnes of catch and USD 78 billion in fishing revenue between 2005 and 2010 [@PalaciosAbrantes:2020tv].

The effectiveness of fisheries management for transboundary stocks is challenged by the shifts in marine stocks distribution under climate change [@Pinsky:2018cb; @Pinsky:2014by]. In many cases, catch or fishing effort quotas for transboundary stocks are based on historical records [@Baudron:2020dj] and do not necessarily consider the biogeography of the species [@FredstonHermann:2018kp], nor the effects of climate change on the fish stocks and fisheries [@PalaciosAbrantes:2020tv; @Sumby:2021gw]. Misalignment between fisheries resources allocation and species’ distributional shifts have previously resulted in unsustainable harvest and international disputes [@Miller:2013iv; @Song:2017va; @Spijkers:2017ij], patterns that are expected to be exacerbated by intensifying climate change [@Pinsky:2018cb]. With future shifts in species distributions under climate change, the challenges for the fisheries management of transboundary stocks will continue to increase during the $21^{st}$ century. In order to prepare ocean governance to meet the resource management challenges arising from shifting transboundary stocks, it is important to constrain when climate change will affect the dynamics of transboundary stocks and the intensity of the resulting impacts [@Pinsky:2018cb; @PalaciosAbrantes:2020tv; @Link:2010ei].

The timing and intensity of climate impacts on managing transboundary fish stocks can be quantitatively examined through the concepts of ‘time of emergence’ (ToE) and ‘threat point’ (TP). Time of emergence is defined as the moment in time when a signal (e.g., future anthropogenic trend) emerges from the background noise of natural variability (e.g., historical natural variation) [@Hawkins:2012hc] and has been applied to multiple oceanic physical and biogeochemical variables [@Rodgers:2015jq; @Keller:2014jv; @Frolicher:2016dg; @Cheung:2020ja; @Schlunegger:2020gz; @Schlunegger:2019el; @Henson:2017ev]. The premise behind the time of emergence is that we can only be confident that a significant change has been detected when the signal of anthropogenic climate change is larger than the background noise of natural climate variability [@Hawkins:2012hc]. The concept of threat point comes from game theory and is defined as the minimum payoff a player is willing to receive to cooperate in a game theoretic model [@Nash:1953vr]. Game theory has been widely used to investigate and manage transboundary stocks [@Sumaila:2013hv], including the impacts of climate change on the economics of transboundary species between Canada and the United States [@Sumaila:2020tp].

Here, we estimate the time of emergence and threat point of the share distribution of transboundary stocks worldwide. We employ a species distribution model driven by output from a comprehensive Earth System Model with ten ensemble members to project the distribution of 663 transboundary species [@PalaciosAbrantes:2020tv] across 280 EEZs of 198 coastal countries under a high greenhouse gas emissions scenario RCP8.5 (see Methods). We treated each species in an EEZ as a single stock due to the lack of more spatially resolved data to delineate the boundary of a population [@PalaciosAbrantes:2020tv] and only considered shared stocks between neighboring EEZs (i.e., transboundary stocks that exclude the high seas). We find that many transboundary stocks have already ‘emerged’ from historical natural variability and exceeded their share threat point in most neighboring EEZs, posing a threat to their sustainability and the resilience of international treaties.

## Materials and methods

### Databases and species selection

This analysis includes 633 exploited marine transboundary species previously identified to account for 80% of the catch taken from the world’s EEZs between 2005 and 2014 [@PalaciosAbrantes:2020tv]. Each time a species was shared by a pair of neighbouring EEZs it was considered a transboundary stock [@PalaciosAbrantes:2020tv]; @Teh:2015gd], resulting in a total of 9,132 transboundary stocks in this analysis. Using the transboundary species Atlantic cod (*Gadus morhua*) as an example, and based on the definition of transboundary stock in this study, the United States and Canada share a stock of Atlantic cod, and Canada and Greenland share another stock of Atlantic cod, but the United States and Greenland do not share a stock of this transboundary species. We defined the boundaries of the world’s EEZs using the Sea Around Us spatial division (updated 1 July 2015, available from http://www.seaaroundus.org), noting that it subdivides the EEZs of 198 coastal states into 280 regions (Figure \@ref(fig:toeFigureS1)) including island territories. We determined the intersections between polygons using the R-statistical software package sf [@PackagesfSimpleF:2018vp]. Each EEZ was categorized by geopolitical region according to the United Nations (https://population.un.org/wpp/DefinitionOfRegions/) and biome [@Reygondeau:2019uh]. The habitat association of each species was determined following the classification in FishBase for fish species (http://www.fishbase.org) and SeaLifeBase for invertebrates (http://www.fishbase.org). For each stock and EEZ, we used the Sea Around Us data to estimate catch and fishing revenue from fishing activities within global EEZs. We report both average catch and revenue for the last decade (2005-2014) [@Zeller2016; @Tai2017; @Sumaila:2015uc]. The monthly average US Consumer Price Index (CPI) according to the U.S. Bureau of Labor Statistic (https://www.bls.gov/cpi/) was used to standardize all monetary values to 2019 real USD.

### Projecting species distributions under climate change

We projected the distribution of marine species using a dynamic bioclimatic envelope model (hereafter called DBEM) [@Cheung:2010dt; @Cheung:2016jd]. The DBEM represents species’ physiology, habitat suitability, depth and latitudinal ranges, and spatial population dynamics as well as preferences for sea temperature, salinity, oxygen content, sea ice extent (for polar species) and bathymetry. For pelagic species, the model uses environmental variables at the surface whereas demersal species distribution is driven by ocean bottom variables . DBEM then estimates species abundance and maximum catch potential (a proxy of maximum sustainable yield) over a $0.5^{\circ}$ x $0.5^{\circ}$ grid (see Cheung et al. 2010 and 2016 for model details). Importantly, DBEM is able to project catches for the world’s large marine ecosystems and changes in catches by Exclusive Economic Zones that are consistent with observational-based estimates of catch from 1950 to 2016 [@Cheung:2016jd], Cheung et al. *in revision*].

We forced the DBEM with simulated ocean conditions from a ten-ensemble member simulation of the Geophysical Fluid Dynamics Laboratory Earth system model (GFDL-ESM2M) to project the distributions of the 633 species from 1951 to 2100 [@Dunne:2012bo; @Dunne:2013db; @Rodgers:2015jq]. The GFDL-ESM2M is run under historical forcing until 2005 and follows the high greenhouse gas emissions scenario, the Representative Concentration Pathway 8.5 (RCP 8.5) over the 2006-2100 period [@Riahi:2011dk]. Because the main approach of this paper relies on understanding the spatial and temporal variation of a stocks’ distribution, we have to understand distribution variability during both the historical and the future periods, to infer differences between time frames. We do this by following an ensemble member approach where each member constitutes a different realization of the Earth system variability condition [@Rodgers:2015jq; @Frolicher:2009ge]. Thus, for our experiment, each of the ten GFDL-ESM2M ensembles were started from infinitesimally small differences in Earth system initial conditions in year 1950 resulting in a unique atmosphere and ocean state at each point in time after about three years for surface waters and eight years for subsurface waters [@Frolicher:2020cc]. By design, variations among ensemble members are then solely due to natural variability. This approach allows us to estimate the natural internal variability (e.g., background noise) and isolate the forced anthropogenic climate signal of a stock’s distribution by averaging the secular trend over all ten ensemble members.

### Calculating an index of transboundary range shift

I developed an index to evaluate range shifts in transboundary stocks under climate change (TI). This index is based on the shifts in the distribution centroid of a transboundary stock relative to the centroid of the neighboring EEZs that share this stock (Figure \@ref(fig:toeFigureS2)). The centroid of a transboundary stock was determined by the average (µ) latitude ($lat_{ts}$) and longitude ($lon_{ts}$) across the grid cells with the highest abundance within the neighboring EEZs sharing the stock e.g., sockeye salmon (*Oncorhynchus nerka*) that is transboundary between the United States and Canada. Therefore,

<!-- $lat_{ts} = \mu (lat_{per})$ -->

<!-- $lon_{ts} = \mu (lon_{per})$ -->

\begin{equation}
\begin{split}
\label{eq:one}
lat_{ts} = \mu (lat_{per}) \\
lon_{ts} = \mu (lon_{per})
\end{split}
\end{equation}

where $lat_{per}$ and $lon_{per}$ are the latitudes and longitudes of the grid cells holding the $per^{th}$ percentile of the projected transboundary stock abundance. To focus on areas where transboundary stocks are more abundant and fishing activities are more likely to take place, we included grid cells where the projected stock abundance within the neighbouring EEZs sharing the stock was above the  95th percentile (e.g., per = 95%). We ran a sensitivity analysis using a subset of species (n = 34) for all EEZs (per = 20th, 50th and $90^{th}$ percentile) and examined the effects of different thresholds on the calculated index value (Figure \@ref(fig:toeFigureS3)). The centroid of each EEZ was estimated using the st package in R (Figure \@ref(fig:toeFigureS1)). For each ensemble member, neighboring EEZs and transboundary stock, the distance between centroids was computed assuming the Earth is a perfect sphere and ignoring geographic barriers using the geosphere package in R;

<!-- $$D_{ens} = acos(sin(lat_{eez})*sin(lat_{ts}) + cos(lat_{eez})*cos(lat_{ts})*cos(lon_{eez}-lon_{ts}))$$ -->

\begin{equation}
\label{eq:two}
D_{ens} = acos(sin(lat_{eez})*sin(lat_{ts}) + cos(lat_{eez})*cos(lat_{ts})*cos(lon_{eez}-lon_{ts}))
\end{equation}

where $lat_{eez}$ and $lat_{ts}$ are the latitudes of the EEZ and transboundary stock centroids, respectively and $lon_{eez}$ and $lon_{ts}$ are the respective longitudes. Then, for each year we calculated the transboundary index (TI) as follows:

<!-- $$TI =(\frac{D_{A,t}}{sd(D_{A,th})} - \frac{D_{B,t}}{sd(D_{B,th)}})^2$$ -->

\begin{equation}
\label{eq:three}
TI =(\frac{D_{A,t}}{sd(D_{A,th})} - \frac{D_{B,t}}{sd(D_{B,th)}})^2
\end{equation}

where $D_A$ and $D_B$ represent the distance between the distribution centroids of a species and the centroid of the pair of neighbouring EEZs (A and B) sharing the species for each time step from 2006 to 2100 (*t*); and *sd* is the standard deviation of the historical (*th*, 1951 - 2005) centroid distribution for $D_A$ and $D_B$. Thus, a higher TI means that a shared stock becomes more or less abundant relative to its neighbouring EEZs that is beyond what the fisheries experienced historically. From the perspective of a state’s fisheries, higher TI indicates an increase in the challenges to effective management and sustainability of the shared stock’ fisheries. In contrast, a smaller TI suggests that the sharing of the stocks between neighbouring EEZs is becoming relatively more stable; a condition that favors effective management and sustainability of the shared stocks’ fisheries. The index was smoothed to a 10-year average to reduce interannual variability.

#### Calculating the time of emergence of the transboundary index

Knowing the point in time (e.g., year) at which the distribution of a shared stock will diverge from its natural internal variability is important in informing the lead-time to which climate adaptation needs to be implemented [@Link:2010ei]. Here, we define time of emergence of the tranboundary index (ETI) as the time when the stock shared between neighbouring EEZs will exceed the historical natural internal variability as follows:

<!-- $EmergenceTIndex = \frac{\mu_{TI[t]}}{SD_{TI_{th}}}$ -->
<!-- $ToE = t when Emergence index[t] > 1$ -->

\begin{equation}
\begin{split}
\label{eq:four}
ETI = \frac{\mu_{TI[t]}}{SD_{TI_{th}}} \\
ToE = t when Emergence index[t] > 1 s.d.
\end{split}
\end{equation}

where $\mu_{TI}$ is the yearly mean TI across all ten ensemble members in year *yr* and $SD_{TI}$ is the standard deviation of the smoothed (10-year moving average) TI across the ensemble members for the historical reference period (1951 - 2005). We set a threshold of TI above which it was consider a shared stock to have emerged from the historical internal variability of the TI. The threshold was set as one standard deviation of the variability ($SD_{TI}$) (i.e., Equation \@ref(eq:four), representing a probability of 68% that the transboundary index has emerged from historical variability). We tested the sensitivity of the calculated emergence index to  a higher emergence threshold of two times the standard deviation (i.e., representing a probability of 95% that the index has emerged).

<!-- #### Computing a transboundary fisheries vulnerability index (VI) -->

<!-- We created a country-specific vulnerability index (VI) to examine the risk of impacts of species’ range shifts on transboundary fisheries management. The VI reflects both the rate of change of shared stocks’ distribution (indicated by the time-of-emergence of the transboundary index) and the importance of the shifting species to a country’s total catch and revenue from the species. The VI is calculated from: -->

<!-- $$ index' = (P_c + P_v)/ToE $$ -->

<!-- \begin{equation} -->
<!-- \label{eq:five} -->
<!-- VI' = \frac{(P_c + P_v)}{ToE} -->
<!-- \end{equation} -->

<!-- where $P_c$ and $P_v$ indicate the proportion accounted for by shared species in total catch and value generated from transboundary fisheries within a country’s EEZ, respectively. Here, the proportion of total catch and value are represented as the average of 2005 to 2014 within individual EEZs according to the Sea Around Us. The VI was then normalized to a 0 (less vulnerable) to 100 (more vulnerable) scale. -->

Estimating the threat point of transboundary stock share

In game theory, cooperation over a transboundary resource will more likely result in a better overall outcome for the sharing parties than those operating on a non-cooperative basis [@Munro:1979dg]. However, the benefits that each player gets from a cooperative strategy must be above a minimum payoff i.e., “threat point” [@Sumaila:2013hv]. Here, we define the threat point as the minimum required proportion of a transboundary stock present within an EEZ for a country to engage in cooperative management with their sharing neighbor [@PalaciosAbrantes:2020vt; @Sumaila:2020tp]. Any proportion below the defined threat point of a country would result in a deflection from the joint management of the transboundary stock.

First we estimated the stock share ratio (SSR) of each transboundary stock shared by neighbouring nations between 1951 and 2100. We did this by aggregating the number of $0.5^{\circ}$ x $0.5^{\circ}$ grid cells in which the stock is present across neighboring EEZs, and then calculated the proportion of the stock held within each EEZ [@PalaciosAbrantes:2020tv]. Second, we average the calculated proportion of stock within an EEZ held into three time periods to reduce the effects of inter-annual climate variability. The first time period ($SSR_{th}$) spans from 1951 to 2005 and matches the historical period in the GFDLESM-2M simulations, over which the model was forced with observation-based greenhouse gas, aerosol and natural external forcing [@Dunne:2012bo; @Dunne:2013db]. It is assumed as the historic baseline of shared distributions for each stock. We then selected two future periods; the early $21^{st}$ century as the average of 2021 to 2040 ($SR{te}$), and the mid $21^{st}$ century as the average of 2041 to 2060 ($SSR{tm}$). We chose these time periods to match the challenges of achieving fisheries-related UN-SDGs (6) such as SDG 14.4 (end overfishing), SDG 2.4 (ensure sustainable food production systems) or SDG 1.2 (poverty reduction), to be achieved by 2030 [@Singh:2017ds]. The analysis was replicated for projected stocks distributions from each of the ten ensemble members and results were averaged across ensemble members. Third, we defined a threat point for each EEZ’s stock as $SSR_{th} \pm \sigma$, where $\sigma$ is the standard deviation of $SSR_{th}$. This way, a change of SSR beyond an EEZ’s threat point happens when the future SSR exceeds one standard deviation of the historical variations of the SSR i.e when $SSR_{et} \geqq (SSR_{th} + \sigma)$ or $SSR_{e,m} \leqq (SSR_{th} - \sigma)$. Finally, we estimated the percentage change in the SSR ($\Delta{SSR}$) of each future time period ($SSR_f$) relative to the historic time period ($SSR_{th}$) for each stock whose share ratio overpassed the threat point following Palacios-Abrantes et al. (*in press*),

<!-- $$ \Delta{SSR_{e,m}} = \frac{(SSR_{f}-SSR_{th})}{SSR_{th}}* 100 $$ -->

\begin{equation}
\label{eq:five}
\Delta{SSR_{e,s}} = \frac{(SSR_{f}-SSR_{th})}{SSR_{th}}* 100
\end{equation}

### Statistical analysis

The time of emergence results were tested for normality (e.g. skewness = 1.02, kurtosis = 2.83) and performed a couple of non-parametric Kruskal–Wallis test by ranks [@Hollander:2013wz] to investigate geopolitical and ecological differences in the ToE of transboundary species. Specifically, we tested if the habitat association of transboundary species and the geographic location of EEZs would have any effect on the time of emergence of the transboundary index of transboundary stocks. For both cases, our null hypothesis was that there were no significant differences in the time of emergence across habitat association nor EEZs.
<!-- Moreover, we implemented a multiple linear regression analysis to a series to explain the ToE of each species in each EEZ. Specifically, we looked at habitat associations, catch trends, geopolitical and biome location and ToE of both bottom and sea surface temperature (See *Databases and species selection*). We hypothesize that all of these factors would have significant influence in the ToE of transboundary species. -->
All analyses were run using the statistical software R version 3.5.2 (2018-12-20; Eggshell Igloo) with the packages data.table [@Packagedatatable:2019uh], janitor [@PackagejanitorSim:2018ur], rfishbase [@PackagerfishbaseR:2019th], sf [@PackagesfSimpleF:2018vp], sp [@PackagespClasses:2019tn], tidiverse [@PackagetidyverseE:2017vq], tidytext [@PackagetidytextTe:2019vk], gmt [@PackagegmtInterfa:2017vh] and zoo [@PackagezooSInfr:2019tm], ggrepel [@PackageggrepelAut:2020to], zeallot [@PackagezeallotMul:2018wy], viridis [@PackageviridisDef:2018us], cowplot [@CowplotStreamlined:2019wt], moments [@PackagemomensMome:2015uh] and pgirmess [@PackagepgirmessSp:2018um]. All code is available at https://github.com/jepa/TransEmergence

## Results

```{r Load_Results, eval = TRUE,  echo = FALSE, include = FALSE}

prop_r <- read_csv("/Volumes/HALL2000/Data/TransEmergence/Results/proportion_change_2005.csv") %>% 
  gather("time_step","change",per_early,per_mid)

Time of emergence of transboundary stocks

The average time of emergence of the transboundary index across the emerging stocks in all EEZs analyzed is projected to be 2029 $\pm$ 26 years (Figure \@ref(fig:toeFigure1)). In total, the projected transboundary index of 5,745 stocks (63% of the studied stocks) will emerge from their historical variability between 2006 and 2100. Fish stocks with transboundary index that first emerged from their historical variability in each EEZ was projected to be as early as 2006 and the last one in 2100. About 55% (n = 3,154) of those emerging stocks are projected to have their transboundary index emerged between 2006 and 2020 (e.g. Peruvean anchoveta - *Engraulis ringens* - shared by Chile and Peru emerged in 2010). The number of EEZs with emerging stocks is projected to increase steeply from 87% in 2006 to 96% in 2020. After the mid-2020s, the rate of increase in the number EEZs slow down to an average of 0.7% per year. By 2081, all of the EEZs analyzed here will have at least one emerging stock (Figure \@ref(fig:toeFigure1)). Moreover, 96% of the world’s EEZs saw at least one stock having their distribution shifts between 2006 and 2020 and around one fifth (19%) of the EEZs have an average time of emergence of the transboundary index across stocks prior to 2020. Shifts in stocks’ distributions will continue to exceed historical limits steadily towards the end of the $21^{st}$ century (Figure \@ref(fig:toeFigure1)). We based our transboundary index on a transboundary stock’s centroid defined as the average grid cells that present the higher stock’s abundance (see Methods - Calculating an index of transboundary range shift). Results from the sensitivity analysis suggest that our method of estimating the time of emergence of the transboundary index is robust to the abundance level (e.g. the top 90%) of the stock (Figure \@ref(fig:toeFigureS3)).

```{r toeFigure1, eval = T,  echo = F, fig.width = 9, fig.height=4, fig.pos= "H", fig.cap =  "Time of emergence by Exclusive Economic Zone (EEZ) and transboundary stock."}

@@

The median time of emergence of the transboundary index of stocks varies significantly according to the geographic region of the neighboring EEZs (Kruskal-Wallis, $X^2$ = 287.23, DF = 93, p < 0.001; Figure \@ref(fig:toeFigure2)A). Overall, most tropical EEZs will see an earlier time of emergence with the EEZs of Latin America and the Caribbean and Polynesia having significantly earlier times of emergence (p < 0.05; see Table \@ref(tab:toeTableS1)) for test statistics) than almost all other regions e.g., Ecuador (ToE = 2013) and Papa New Guinea (ToE = 2018) (Figure \@ref(fig:toeFigure2)B). In contrast, EEZs located in template regions like northern Europe and eastern Asia will have a significantly later time of emergence (p < 0.05; see Table \@ref(tab:toeTableS1 for test statistics) than the rest of the world, e.g., Ireland (ToE = 2067) and North Korea (ToE = 2058). However, notable exceptions of tropical EEZs with a later average time of emergence across stocks include Pacific EEZ of Honduras (ToE = 2071) and Panama (ToE = 2070). Moreover, despite being outside the tropics, Belgium (ToE = 2015) and Norway (ToE = 2019) will have relatively early average time of emergence (Figure \@ref(fig:toeFigure2)A). Only in a few cases, the transboundary index of a stock will not emerge between 2006 and 2100. In the Arctic EEZs of Canada and the US, the present study included three stocks: capelin (*Mallotus villosus*), saffron cod (*Eleginus gracilis*), and Pacific herring (*Clupea pallasii*). In the case of Brazil, there were four transboundary stocks included, all shared with Uruguay [@PalaciosAbrantes:2020tv]: Argentine menhaden (*Brevoortia pectinata*), Argentine shortfin squid (*Illex argentinus*), Argentine shrimp (*Pleoticus muelleri*), and pink shrimp (*Penaeus paulensis*). In all cases, our study suggest that the transboundary index of these stocks will not emerge before 2100 relative to the historic (1951-2005) distribution (Figure \@ref(fig:toeFigure2)A)

>>>>

---
title: Identifying timing of shifts in transboundary fish stocks to support climate resilient fisheries management

author: 'Authors: Juliano Palacios-Abrantes^1*^, Thomas Frölicher^2,3^, Gabriel Reygondeau^1^, U. Rashid Sumaila^1^, Alessandro Tagliabue^4^, Colette Wabnitz^1,5^, and William W.L. Cheung^1^'
date: 'Affiliations: ^1^Institute for the Oceans and Fisheries, University of British Columbia, Vancouver, Canada. ^2^Climate and Environmental Physics, Physics Institute, University of Bern. ^3^Oeschger Centre for Climate Change Research, University of Bern, Switzerland. ^4^School of Environmental Sciences, University of Liverpool. ^5^Stanford Center for Ocean Solutions, Stanford University, Stanford, United States'
output:
  bookdown::word_document2
reference_docx:  ../References/manuscript_template.docx
  # pdf_document: default
bibliography: ../References/ToE_Ref.bib
csl: ../References/science.csl
header-includes:
- \usepackage{amsmath}
- \usepackage{float} # Use the 'float' package
- \floatplacement{figure}{H} # Make every figure with caption = h
---

@@

-   corresponding author: [j.palacios\@oceans.ubc.ca](mailto:j.palacios@oceans.ubc.ca){.email}

**Abstract**

@@

Climate change is shifting the distribution of transboundary fish stocks between neighbouring countries. The time-frame of such shifts determine the extent of climate challenges to international fisheries governance. Here, using global climate-fisheries simulation modelling, we show that, by 2036, 68% of the world’s transboundary stocks are projected to shift their distributions significantly across international boundaries. By 2100, such shifts will occur in all countries under a ‘no mitigation’ scenario. Hotspots where transboundary stock shifts are earlier and larger coincide with countries that are highly dependent on fisheries for income and livelihood. The early occurrences of transboundary shifts highlight the urgency to re-formulate international fisheries policies to avoid disputes in fish stock sharing. Such bi-/multi-lateral discussions should start now as lead-time for policy changes is long.

**One-Sentence Summary**

Early range shifts in transboundary fish stocks challenge their management under climate change

**Main Text.**

Over the last century, human activities have altered the physical and biogeochemical conditions of the ocean, resulting in warmer, more acidic and less oxygenated waters, [@IPCC:2019tn]. Marine species' distributions reflect species' preferences for discrete environmental conditions [@Hutchinson:1957nt; @Nelson:2016kk]. As a result of climate change, many marine species have shifted their distributions towards higher latitude, deeper water or following local temperature gradients to track and stay within their optimal environmental 'niche'[@Poloczanska:2016kk; @Barange:2014kr]. The biogeography of marine species is projected to continue to shift as ocean conditions change through the 21st century [@Cheung:2010dt; @Barange:2014kr], impacting fisheries production and economies [@Sumaila:2019ew], and compromising our capacity to reach international sustainability goals within the 2030 Agenda such as Sustainable Development Goal 14 - life below water [@UnitedNations2018; @Singh:2017ds; @Pecl:2017hu]. The projected risks of impacts can be reduced by improving the effectiveness of current fisheries management [@Cheung:2018dh], including for species that cross international borders, i.e., shared stocks [@Pinsky:2018cb].

The concept of shared stocks was developed following the ratification of the United Nations Convention on the Law of the Sea (UNCLOS) and the claiming of Exclusive Economic Zones (EEZs) by Coastal States [@UnitedNations:1986tl]. As defined by the United Nations' Food and Agriculture Organization (FAO), shared stocks can be classified into four non-exclusive categories: (*i*) transboundary stocks, which cross neighboring EEZs; (*ii*) straddling stocks that, in addition to neighboring EEZs, also visit the adjacent high seas; (*iii*) highly migratory stocks, mainly tunas and bill-fishes, that migrate across vast oceanic regions including both the high seas and EEZs; and finally (*iv*) discrete stocks that are only present on the high seas [@Munro:2004th]. This study focuses on transboundary stocks exploited by fisheries operating within EEZs. Countries are responsible for the management of stocks within their EEZs, under UNCLOS, and are encouraged to cooperate when stocks are shared [@UnitedNations:1986tl]. A recent study estimates that, globally, there are 633 transboundary fish species representing 67% of identified exploited taxa. These species yielded an annual average of 48 million tonnes of catch and USD 78 billion in fishing revenue between 2005 and 2010 [@PalaciosAbrantes:2020tv].

The effectiveness of fisheries management for transboundary species, and thus reaching the 2030 Agenda [@Pecl:2017hu], is challenged by shifts in stocks' distribution under climate change [@Pinsky:2018cb; @Pinsky:2014by; @Pinsky:2020jx; @Oremus:2020id]. In most cases, catch or fishing effort quotas for transboundary stocks are based on historical records [@Baudron:2020dj] and do not necessarily consider the full distribution range of the stock [@FredstonHermann:2018kp], nor the effects of climate change on fish stocks and associated fisheries [@PalaciosAbrantes:2020tv; @Sumby:2021gw]. While many transboundary stocks experience natural seasonal spatial variation (e.g., migration, reproductive cycle), misalignment between fisheries resources allocation and distributional shifts beyond such natural variation have previously resulted in unsustainable harvest and international disputes [@Miller:2013iv; @Song:2017iua; @Spijkers:2017ij; @Crespo:2020cs]. Such patterns are expected to be exacerbated by continuing climate change [@Pinsky:2018cb; @Sumaila:2011bq]. The importance of transboundary stocks for both catch and fisheries revenue [@PalaciosAbrantes:2020tv] highlights their critical role in achieving the SDGs. Understanding when climate change will affect the sharing dynamics of transboundary stocks and the intensity of the resulting impacts is important to developing climate resilient international ocean governance and achieving the 2030 Agenda [@Pinsky:2018cb; @Link:2010ei; @Sumaila:2020tp].

Here, we estimate the year in which 663 transboundary species (9,132 stocks) are projected to shift their shared distribution beyond historical variation and the intensity of said shift.  We treat each species in an EEZ as a single stock due to the lack of more spatially resolved data to delineate the boundary of a population [@PalaciosAbrantes:2020tv] and only consider shared stocks between neighboring EEZs (i.e., excluding the high seas). We employ a mechanistic population dynamic model driven by output from a comprehensive Earth System Model with ten ensemble members to project transboundary stocks' distribution across 280 EEZs of 198 coastal countries under a high greenhouse gas emissions scenario (RCP8.5; Methods). We find that the shared distribution of many transboundary stocks has already shifted beyond historical natural variability. Moreover, by 2030, the shared catch proportion of such stocks will exceed historical levels in most neighboring EEZs, posing a threat to the sustainability of associated fisheries and the resilience of international treaties that manage their shared exploitation.

**Identifying range shifts in the shared distribution of transboundary stocks**

We calculated the time of emergence of the transboundary index to identify shifts in the shared distribution of transboundary stocks from 2006 to 2100 under a high emission climate change scenario. Our results suggest that 4,119 transboundary stocks will experience a range shift beyond historical variability as early as 2006 and up to 2100 (Fig. \@ref(fig:toeFigure1)A). This corresponds to 68% of the studied stocks (Fig. \@ref(fig:toeFigure1)C). On average, the shift in shared distribution of these stocks in all EEZs analyzed is projected to be 2036 $\pm$ 28 years. Furthermore, 96% of the world’s EEZs will experience at least one transboundary stock shift between 2006 and 2100 with around 80% presenting an early distributional shift across stocks (Fig. \@ref(fig:toeFigure1)C).

```{r toeFigure1, eval = T,  echo = F, fig.width = 9, fig.height=4, fig.pos= "H", fig.cap =  "Timeline of shifts in transboundary stock shared distribution. Dashed lines represent the year by which countries have committed to reach the full implementation of the 2030 Agenda, as well as an early and late change in range shift, respectively."}

@@

The median year in which transboundary stocks experience a range shift varies significantly according to the geographic region of the neighboring EEZs (Kruskal-Wallis, $X^2$ = 242.11, DF = 93, *p* < 0.001; Figure \@ref(fig:toeFigure2)A). Overall, most tropical EEZs will see earlier shifts, with the EEZs of Latin America, the Caribbean and Polynesia being significantly earlier (*p* < 0.05; see Table \@ref(tab:toeTableS2) for test statistics) than almost all other regions (Fig. \@ref(fig:toeFigure2)B). In contrast, EEZs located in temperate regions like northern Europe and eastern Asia are projected to experience significantly later range shifts (Table \@ref(tab:toeTableS2)). Stocks in some countries, like Brazil or New Zealand, are not projected to experience range shifts beyond natural variability (e.g., the transboundary index of a stock did not emerge from historical variability under a 95% confidence interval) (Fig. \@ref(fig:toeFigure2)A). On average, shifting transboundary stocks represented 23% $\pm$ 22% of yearly fishing revenue from fisheries targeting such stocks within global EEZs between 2004 and 2010 (Fig. \@ref(fig:toeFigure2)A). However, large variation exists, with shifting stocks representing less than 1% of fishing revenue in some countries (e.g., Ireland) and over 90% in others (e.g., Marshall Islands). Moreover, in some EEZs, while few transboundary stocks are projected to shift above historical variability, they still represent a large proportion of revenue derived from fisheries within that EEZ (e.g., Peru).

```{r toeFigure2, eval = T,  echo = F, fig.width  = 13,fig.pos= "H", fig.height =11, fig.cap =  "Year of change in the shared distribution of 4,119 transboundary stocks. A) Land polygon shows the contribution of shifting stocks to the country’s total fishing revenue from transboundary stocks. Exclusive Economic Zone polygons represent average range shifts per EEZ. EEZs with no time of emergence between 2006 and 2100 are represented in aqua color. B) Year of shared distribution shifts by region according to United Nations sub-regions. Points represent the median year of range shift. Horizontal dashed line represents the 2030 Agenda. N = North, S = South, W = West and E = East. Ltn. Ame. and the Car. = Latin America and the Caribbean. Aus and New Z. = Australia and New Zealand"}
